---
title: "p8105_hw2_qra2000"
output: github_document
date: "2023-09-26"
---

```{r load_libraries}
library(tidyverse)
library(readxl)
```

## Problem 1

First, clean the data in `pols-month.csv`. Use `separate()` to break up the variable `mon` into integer variables `year`, `month`, and `day`; replace `month number` with `month name`; create a president variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop`; and remove the `day` variable.

```{r clean_538_pols}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pols = 
  read_csv("./data/fivethirtyeight_datasets/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez"))
```

Second, clean the data in `snp.csv` using a similar process to the above. For consistency across datasets, arrange according to `year` and `month`, and organize so that `year` and `month` are the leading columns.

```{r clean_538_snp}
snp = 
  read_csv("./data/fivethirtyeight_datasets/snp.csv") |>
  separate(date, into = c("month", "day", "year"), convert = TRUE) |>
  arrange(year, month) |>
  mutate(month = month.name[month]) |>
  select(year, month, close) 
```

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r clean_538_unemp}
unemployment = 
  read_csv("./data/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```

Join the datasets by merging snp into pols, and merging unemployment into the result.

```{r merge_538}
data_538 = 
  left_join(pols, snp) |>
  left_join(x = _, y = unemployment)

str(data_538)
```

## Problem 2

Read and clean the Mr. Trash Wheel sheet:
* specify the sheet in the Excel file and to omit non-data entries (rows with notes / 
* figures; columns containing notes) using arguments in read_excel
* use reasonable variable names
* omit rows that do not include dumpster-specific data











